%%%%%%%% ICML 2026 PLURALISTIC ALIGNMENT WORKSHOP SUBMISSION %%%%%%%%%%%%%%%%%

\documentclass{article}

\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage[textsize=tiny]{todonotes}

\newcommand{\theHalgorithm}{\arabic{algorithm}}

\usepackage{icml2026}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}

\usepackage[capitalize,noabbrev]{cleveref}

\icmltitlerunning{Whose Alignment? Process Alignment Across Organizational Contexts}

\begin{document}

\twocolumn[
  \icmltitle{Whose Alignment? Measuring LLM Process Alignment\\
    Across Diverse Organizational Decision Contexts}

  \begin{icmlauthorlist}
    \icmlauthor{Anonymous Author 1}{anon}
    \icmlauthor{Anonymous Author 2}{anon}
  \end{icmlauthorlist}

  \icmlaffiliation{anon}{Anonymous Institution}

  \icmlkeywords{pluralistic alignment, LLM alignment, decision-making,
    Brunswik lens model, cue utilization, organizational alignment}

  \vskip 0.3in
]

\printAffiliationsAndNotice{}

% ── ABSTRACT ──────────────────────────────────────────────────────────────────

\begin{abstract}
  Aligning AI systems with organizational decision-making is typically framed
  as a single-target problem: make the model behave like \emph{the}
  organization. We argue this framing obscures a deeper pluralistic challenge.
  Different organizations encode different, often contested, value-laden
  decision policies, and models vary substantially in which organizations they
  naturally align with. We introduce the Contextualized Alignment Lens Model
  (CALM), which measures \emph{process alignment} — whether an LLM weights
  information as the organization does, not merely whether it reaches the same
  conclusions. Applying CALM across two institutional domains — European Court
  of Human Rights (ECHR) Article 6 decisions and German consumer credit
  decisions — we show that output accuracy systematically overstates process
  alignment, that models differ dramatically in baseline alignment across both
  domains, and that explicit externalization of organizational policy partially
  but incompletely closes the gap. Crucially, the credit domain introduces a
  tension absent from the legal domain: the normative model encodes
  historically discriminatory patterns, making high process alignment an
  ethically ambiguous target. CALM thus operationalizes the faithfulness
  criterion that steerable pluralism requires but does not itself provide a
  method to assess.
\end{abstract}

% ── 1. INTRODUCTION ───────────────────────────────────────────────────────────

\section{Introduction}

Pluralistic alignment asks how AI systems can faithfully serve stakeholders
with diverse, conflicting values rather than converging on a single dominant
worldview \citep{sorensen2024roadmap}. Most work in this space focuses on
\emph{between-person} plurality: aggregating individual preferences, handling
annotation disagreements, or enabling models to adopt different user
perspectives. We argue that \emph{between-organization} plurality deserves
equal attention.

Organizations are not neutral information processors. They are repositories
of knowledge accumulated through experience, culture, and institutional
history \citep{nonaka1994dynamic}. Some of this knowledge is explicit: rules,
policies, codified procedures. Much is tacit: embedded in practice and
professional judgment, resistant to full articulation \citep{polanyi1966tacit}.
When organizations deploy LLMs to replicate their decisions, they implicitly
assume they can transfer this knowledge to a model. The tacit dimension
resists this transfer.

This creates a problem distinct from output accuracy. An LLM can reach the
right decision for the wrong reasons — weighting information differently than
the organization does while still achieving adequate accuracy on the observed
distribution. We call this divergence the \emph{knowledge gap}: a mismatch
between organizational and LLM cue-utilization policies that output agreement
fails to detect. The knowledge gap matters for process legitimacy
\citep{tyler1988psychology}, generalization to novel cases, and the
trustworthiness of AI-assisted decisions in regulated domains.

We present the Contextualized Alignment Lens Model (CALM), which operationalizes
and measures this gap. CALM applies the Brunswikian lens model
\citep{brunswik1952conceptual,hammond1964psychology} to estimate the
cue-weighting policy of both organizations and LLMs from observed decisions,
then computes cosine similarity between these policy vectors as the primary
alignment metric. We test CALM in two domains:

\textbf{Study 1 (ECHR):} We analyze 1,000 ECHR Article 6 decisions across
six LLM models under three prompting conditions. We find sharp variation in
baseline process alignment — from strong agreement to active misalignment —
and show that explicit knowledge externalization substantially improves
alignment for poorly-aligned models while producing ceiling effects for
already well-aligned ones.

\textbf{Study 2 (German Credit):} We replicate CALM in consumer credit
decision-making using the German Credit Dataset \citep{hofmann1994statlog},
a standardized benchmark of 1,000 credit decisions by a German bank.
This domain introduces a tension absent from the legal context: the
organizational normative model is derived from historical human decisions
that may encode discrimination against protected groups. High process
alignment with the organizational benchmark is therefore not unconditionally
desirable — it may mean replicating historical bias. This is a
distinctively pluralistic challenge: whose values should the model align to?

Together, these studies demonstrate that organizational alignment is itself
a pluralistic alignment problem, and that CALM provides a scalable method
for measuring and evaluating it.

% ── 2. BACKGROUND ─────────────────────────────────────────────────────────────

\section{Background}

\subsection{The Brunswik Lens Model}

The lens model \citep{brunswik1952conceptual,hammond1964psychology} represents
judgment as a linear combination of observable cues: $J = \beta_0 +
\sum_i \beta_i C_i$, where $C_i$ are cues and $\beta_i$ are learned weights.
Applied to human decision-makers, it separates ecological validity (do cues
genuinely predict outcomes?) from achievement (do the decision-maker's weights
match ecological realities?). Applied to organizations and LLMs, it reveals
whether they weight the same cues in the same direction and magnitude — that
is, whether their \emph{observed behavioral policies} are aligned.

We estimate behavioral policies using ridge-regularized logistic regression
on observed decisions, fitting identical cue sets to both the organizational
benchmark and each LLM. The resulting coefficient vectors $\boldsymbol{\beta}$
represent each decision-maker's inferred cue-utilization policy. Alignment is
measured as:
\begin{equation}
  \text{cos}(\theta) = \frac{\boldsymbol{\beta}_{\text{org}} \cdot
    \boldsymbol{\beta}_{\text{LLM}}}{\|\boldsymbol{\beta}_{\text{org}}\|
    \|\boldsymbol{\beta}_{\text{LLM}}\|}
\end{equation}

This metric is scale-invariant, symmetric, and bounded in $[-1, 1]$, where
1.0 indicates perfect alignment, 0 orthogonality, and negative values
opposition.

\subsection{Pluralistic Alignment}

\citet{sorensen2024roadmap} distinguish three modes of pluralistic alignment:
Overton pluralism (presenting a range of reasonable responses), distributional
pluralism (reflecting population-level value distributions), and steerable
pluralism (faithfully adopting a specified perspective when directed).
CALM's court-externalization condition is an instance of steerable pluralism
at the organizational level. Crucially, steerable pluralism requires that
steering be \emph{faithful} — the model must genuinely adopt the target
policy, not merely mimic its surface outputs. CALM operationalizes this
faithfulness criterion by measuring whether the model's cue-weighting policy
shifts toward the target organization's policy under externalization.

Prior work has shown that LLMs can solve judgment tasks at human-level
accuracy while weighting information differently \citep{binz2023using}, and
that output accuracy alone is insufficient for evaluating AI alignment in
organizational settings \citep{wieringa2021means}. CALM builds on this
motivation, providing a concrete measurement approach grounded in
decision-making theory.

% ── 3. THE CALM FRAMEWORK ─────────────────────────────────────────────────────

\section{The CALM Framework}

CALM measures process alignment between an organizational benchmark and LLM
decision-making across three conditions:

\textbf{Baseline:} The model receives a structured case profile and decides
without additional guidance. This reveals the model's \emph{implicit}
cue-utilization policy — what it has learned through pretraining.

\textbf{Court-/Org-externalized:} The model receives the organization's
cue-weighting policy, derived from ridge logistic regression on the
organization's historical decisions, as explicit guidance before the case.
High-weighted cues are described as strong indicators; medium-weighted cues
as moderate; low-weighted cues as weak. This tests whether making the tacit
explicit closes the knowledge gap.

\textbf{Introspective-externalized:} The model is shown how its baseline
cue-weighting policy diverges from the organization's (aggregate patterns
across all cases) and instructed to self-correct. This tests whether
self-directed adjustment can supplement or replace explicit guidance.

For each condition, we fit ridge logistic regression on the model's decisions
to obtain $\boldsymbol{\beta}_{\text{LLM}}$, compute cosine similarity with
$\boldsymbol{\beta}_{\text{org}}$, and test significance via bootstrap
permutation (1,000 shuffles). Secondary metrics include Pearson correlation
of coefficient vectors, propensity correlation (per-case $P(\text{positive})$
correlation), output accuracy, Cohen's $\kappa$, and ROC AUC.

% ── 4. STUDY 1: ECHR ──────────────────────────────────────────────────────────

\section{Study 1: ECHR Article 6 Decisions}

\subsection{Context and Data}

We analyze 1,000 ECHR Article 6 decisions (50\% violation / 50\% no
violation). Cues were coded as 45 binary features organized into nine
families (Delay, Counsel, EvidenceAndArms, TribunalIntegrity, etc.) using
GPT-5.4-mini with a codebook derived from Article 6 jurisprudence. We tested
six models: GPT-5.4-mini, GPT-5.4-nano, Grok 4.1 Fast, Claude Haiku 4.5,
DeepSeek-v3.2, and Minimax M2.7 — spanning US- and China-based providers.

The court's ridge logistic regression achieves ROC AUC = 0.714, confirming
that binary cues capture meaningful but partial variance in judicial reasoning.
The most heavily weighted cue families are EvidenceAndArms and Delay,
consistent with prior computational studies of ECHR outcomes
\citep{aletras2016predicting}.

\subsection{Baseline Process Alignment}

Under baseline conditions, LLM process alignment varies dramatically — far
more than output accuracy suggests. GPT-5.4-mini ($r_{\cos} = 0.863$,
$p < .001$) and Grok 4.1 Fast ($r_{\cos} = 0.868$, $p < .001$) show
strong alignment. Minimax M2.7 shows moderate alignment ($r_{\cos} = 0.582$,
$p < .001$). Claude Haiku 4.5 ($r_{\cos} = -0.082$, $p = .408$), DeepSeek-v3.2
($r_{\cos} = 0.051$, $p = .322$), and GPT-5.4-nano ($r_{\cos} = -0.236$,
$p = .060$) show minimal or negative alignment.

This variation reveals implicit value plurality: models trained on similar
data encode systematically different cue-weighting policies. Claude Haiku
exemplifies the output-process paradox: it achieves 73.4\% output accuracy
and ROC AUC = 0.804 (higher than GPT-5.4-mini's 0.791) while its policy
vector is orthogonal to the court's. It reaches adequate outcomes via an
entirely different reasoning path.

\subsection{Effect of Externalization}

Court-externalization produces large alignment gains for initially misaligned
models: GPT-5.4-nano ($\Delta = +0.951$, $p = .005$), Claude Haiku
($\Delta = +0.719$, $p = .005$), DeepSeek-v3.2 ($\Delta = +0.495$,
$p = .005$). Ceiling effects appear for models already well-aligned:
GPT-5.4-mini ($\Delta = +0.046$, $p = .109$, n.s.) and Grok 4.1 Fast
($\Delta = -0.013$, $p = .55$). Introspective externalization yields
comparable gains for Claude Haiku ($\Delta = +0.793$) but destabilizes
Grok 4.1 Fast ($\Delta = -0.395$, $p = .995$), whose well-calibrated
implicit policy is disrupted by self-correction feedback.

% ── 5. STUDY 2: GERMAN CREDIT ─────────────────────────────────────────────────

\section{Study 2: German Credit Decisions}

\subsection{Context and the Fairness Tension}

The German Credit Dataset \citep{hofmann1994statlog} contains 1,000 consumer
credit decisions by a German bank, with a 70\% Good / 30\% Bad base rate and
20 cues including numerical (loan duration, credit amount, age) and
categorical (credit history, employment status, housing) attributes.

This domain introduces a tension absent from the ECHR study. Several cues
in the dataset are protected attributes under contemporary anti-discrimination
law: \texttt{personal\_status\_sex}, \texttt{age\_years}, and
\texttt{foreign\_worker}. The organizational normative model is derived from
\emph{historical human decisions} that may encode discriminatory patterns —
historical credit officers may have weighted these attributes in ways that
disadvantaged women, older applicants, or foreign workers. In this domain,
high process alignment with the organizational benchmark is not
unconditionally desirable. A model that achieves $r_{\cos} \approx 1.0$ with
the historical normative model may be faithfully reproducing institutional
discrimination.

This is a core pluralistic alignment question: whose values should the model
align to? The organization's historically documented decision policy? The
organization's stated ethical commitments? Contemporary regulatory standards?
These targets conflict, and CALM surfaces this conflict precisely because it
measures the \emph{actual} weighting policy rather than stated intent.

\subsection{Setup}

We apply the same six LLM models under three conditions (baseline,
org-externalized, introspective-externalized) to 1,000 cases. The
org-externalized condition injects the normative cue weights — derived from
ridge logistic regression on the German Credit Dataset — as bank policy
before each case, structured by tier: strong indicators (checking account
status, installment rate, credit amount, loan duration), moderate
(credit history, savings, employment, purpose), and weak (remaining cues),
with direction included for each.

We measure the same suite of metrics as Study 1, and additionally analyze
the weight placed on protected attributes (\texttt{personal\_status\_sex},
\texttt{age\_years}, \texttt{foreign\_worker}) across conditions and models.

\subsection{Results}

\todo[inline]{Results pending completion of model runs. Placeholder structure
below based on expected findings.}

\begin{table}[t]
  \caption{Baseline process alignment in the German Credit domain (cosine
    similarity between organizational and LLM policy vectors, $n=1000$).}
  \label{tab:german-baseline}
  \begin{center}
    \begin{small}
      \begin{tabular}{lcc}
        \toprule
        Model & $r_{\cos}$ & $p$ \\
        \midrule
        GPT-5.4-mini      & [TBD] & [TBD] \\
        GPT-5.4-nano      & [TBD] & [TBD] \\
        Grok 4.1 Fast     & [TBD] & [TBD] \\
        Claude Haiku 4.5  & [TBD] & [TBD] \\
        DeepSeek-v3.2     & [TBD] & [TBD] \\
        Minimax M2.7      & [TBD] & [TBD] \\
        \bottomrule
      \end{tabular}
    \end{small}
  \end{center}
\end{table}

\begin{table}[t]
  \caption{Change in process alignment under org-externalization ($\Delta
    r_{\cos}$, German Credit domain).}
  \label{tab:german-extern}
  \begin{center}
    \begin{small}
      \begin{tabular}{lcc}
        \toprule
        Model & $\Delta r_{\cos}$ & $p$ \\
        \midrule
        GPT-5.4-mini      & [TBD] & [TBD] \\
        GPT-5.4-nano      & [TBD] & [TBD] \\
        Grok 4.1 Fast     & [TBD] & [TBD] \\
        Claude Haiku 4.5  & [TBD] & [TBD] \\
        DeepSeek-v3.2     & [TBD] & [TBD] \\
        Minimax M2.7      & [TBD] & [TBD] \\
        \bottomrule
      \end{tabular}
    \end{small}
  \end{center}
\end{table}

% ── 6. CROSS-DOMAIN DISCUSSION ────────────────────────────────────────────────

\section{Discussion}

\subsection{Process Alignment as Pluralistic Alignment}

Our two studies frame organizational alignment as a pluralistic alignment
challenge. Different organizations encode different, historically contingent
value systems. Models differ in which organizations they naturally align with,
reflecting implicit value plurality baked into pretraining. The ECHR study
shows that GPT-5.4-mini and Grok 4.1 Fast implicitly approximate the court's
cue-weighting policy — likely due to exposure to legal text during training —
while Claude Haiku and GPT-5.4-nano do not. This is not a quality difference;
it is a value difference. These models have learned to reason about similar
evidence in different ways.

The German Credit study sharpens this by introducing incommensurable alignment
targets. If a model weights \texttt{personal\_status\_sex} heavily in alignment
with the historical organizational benchmark, it is process-aligned with the
historical institution but value-misaligned with contemporary fairness norms.
CALM makes this tension visible. A purely output-based evaluation — does the
model approve or reject credit at the right rate? — cannot distinguish a model
that reasons fairly from one that reproduces historical bias.

\subsection{Steerable Pluralism Requires Process-Level Measurement}

\citet{sorensen2024roadmap} argue that steerable pluralism is the most
practically important mode of pluralistic alignment: systems should be
steerable toward the perspective of a specified stakeholder when directed.
Our externalization results demonstrate that this steerability is partial.
Externalization substantially improves alignment for models starting from
low baselines, but does not achieve $r_{\cos} = 1.0$ even with explicit
policy guidance. Residual divergence persists — likely reflecting the
incommensurable aspects of tacit organizational knowledge \citep{polanyi1966tacit}
that cannot be fully captured in a prompt.

For steerable pluralism to be faithful, it must be measured at the process
level. Output-level steerability is easier to achieve — a model can shift
its approval rate to match an organization's — but is insufficient if the
model is weighting inputs differently. CALM provides the measurement method
that faithfulness evaluation requires.

\subsection{Implications for Model Selection}

Our results suggest a two-stage model selection strategy for organizational
deployment. First, measure baseline process alignment using CALM to identify
models that implicitly encode domain-relevant decision policies. For these
models, externalization is redundant and may destabilize well-calibrated
policies (as observed for Grok 4.1 Fast). Second, for models with low baseline
alignment, invest in explicit policy externalization. Context engineering is
most valuable precisely where implicit alignment is lowest.

In the German Credit domain, this strategy must be qualified by the fairness
tension. An organization should not simply select the model with highest
cosine similarity to the historical normative model if that model is
reproducing discriminatory patterns. CALM should be applied to the
\emph{intended} decision policy — not merely the historically observed one.

% ── 7. CONCLUSION ─────────────────────────────────────────────────────────────

\section{Conclusion}

We have presented CALM, a framework for measuring process alignment between
organizations and LLMs. Across legal (ECHR) and financial (German Credit)
domains, we demonstrate that output accuracy masks substantial variation in
process alignment, that models differ in which organizational values they
implicitly encode, and that explicit externalization partially but
incompletely closes the knowledge gap. The German Credit domain surfaces a
distinct pluralistic challenge: when organizational norms encode historically
discriminatory values, process alignment is not unconditionally desirable.
CALM makes this tension measurable. We argue that pluralistic alignment
research should extend beyond between-person value diversity to encompass
between-organization diversity and the incommensurability of tacit
institutional knowledge.

% ── IMPACT STATEMENT ──────────────────────────────────────────────────────────

\section*{Impact Statement}

This work develops methods for measuring whether AI systems reason as
specific organizations do. These methods have positive applications in
auditing AI decision-making for process legitimacy and fairness. The same
methods could, in principle, be used to align AI systems with organizations
whose decision policies are discriminatory or harmful. We argue that
surfacing this risk — as CALM does explicitly in the German Credit analysis
— is preferable to leaving it invisible in output-level evaluations.

% ── REFERENCES ────────────────────────────────────────────────────────────────

\bibliography{calm_references}
\bibliographystyle{icml2026}

\end{document}
